---
title: 'Causes of Death - Data Sets from Our World In Data'
author: "H. David Shea"
date: 2021-04-21
output:
    github_document: default
---

```{r setup, include = FALSE, warning = FALSE}
library(tidyverse)
library(patchwork)

knitr::opts_chunk$set(
    connection = "secdb", # automatically uses this connection in sql chunks 
    comment = "#>", 
    collapse = TRUE, 
    message = FALSE,
    fig.width = 8,
    fig.asp = ((1 + sqrt(5)) / 2) - 1, # the golden ratio - technically, the b proportion of a+b when a is 1
    out.width = "70%",
    fig.align = "center"
)

causes <- c("Executions", "Meningitis", "Lower_Respiratory_Infections",
            "Intestinal_Infectious_Diseases", "Protein-Energy_Malnutrition", "Terrorism",
            "Cardiovascular_Diseases", "Dementias", "Kidney_Disease", 
            "Respiratory_Diseases", "Liver_Diseases", "Digestive_Diseases", 
            "Hepatitis", "Cancers", "Parkinson_Disease", "Fire", "Malaria", "Drowning", 
            "Homicide", "HIV/AIDS", "Drug_Use_Disorders", "Tuberculosis", 
            "Road_Injuries", "Maternal_Disorders", "Neonatal_Disorders", "Alcohol_Use_Disorders",
            "Natural_Disasters", "Diarrheal_Diseases", "Heat_And_Cold_Exposure", 
            "Nutritional_Deficiencies", "Suicide", "Conflict", "Diabetes", "Poisonings")

causes_of_death <- tibble(read.csv("Data/annual-number-of-deaths-by-cause.csv")) %>%
  arrange(Entity, Year)
colnames(causes_of_death) <- c("Entity", "Code", "Year", all_of(causes))
causes_of_death <- causes_of_death %>% mutate(Executions = as.numeric(Executions))
```

## Data sets

This work references causes of death from the Our World In Data organization.  It reference the specific section of the site [_here_](https://ourworldindata.org/causes-of-death).  

## Causes of Death

The _Global Burden of Disease_ is a major global study on the causes of death and disease published in the medical journal [_The Lancet_](https://www.thelancet.com/gbd). These estimates of the annual number of deaths by cause are shown in this analysis. 

```{r cause_of_death_world, include = TRUE, warning = FALSE, echo = FALSE, out.width = "90%"}
latest_year <- max(causes_of_death$Year)

g <- causes_of_death %>%
  filter(Entity == "World") %>%
  filter(Year == latest_year) %>%
  pivot_longer(all_of(causes), names_to = "Cause", values_to = "Count") %>%
  filter(!is.na(Count)) %>% 
  mutate(
    Count = round(Count / 1000000, 2),
    Cause = reorder(as_factor(str_replace_all(Cause, "_", " ")), Count)
  ) %>% 
  ggplot(aes(Count, Cause)) +
  geom_col(aes(fill = Count)) +
  theme_minimal() +
  theme(legend.position = 'none') +
  scale_fill_distiller(palette = "Dark2") +
  geom_label(aes(label = Count), size = 3, hjust = -0.1, check_overlap = TRUE, 
             label.size = 0.1, label.padding = unit(0.1, "lines")) +
  scale_x_continuous(limits = c(0, 20), breaks = seq(0, 20, by = 5)) +
  labs(
    x = "Millions of Deaths",
    y = ""
  )

g + plot_annotation(
  title = str_c("Number of Deaths by Cause - World - ", latest_year),
  caption = "Source: Our World In Data"
)
```

